---
title: "PHP 2517 HW#3"
author: "Blain Morin"
date: "March 16, 2019"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library ("arm")
library("foreign")
library("readstata13")
library("geepack")
library(dplyr)
library(kableExtra)
library(stargazer)
```

```{r, echo = FALSE, results = 'hide', cache = TRUE, include = FALSE}

### Hogan code

## This section reads in the ctq data and does some pre-processing
ctq = read.dta13("ctq_mlm.dta")

ctq$include = rep(1, nrow(ctq))
for (i in unique(ctq$id)) { 
  nmiss                   = sum(is.na( ctq$Y[(ctq$id == i)] ) )
  ctq$include[ctq$id==i]  = rep( ifelse(nmiss<9, 1, 0), 9 )
}

ctq = ctq[(ctq$include==1),]

ctq = ctq[(ctq$week>4),]
ctq$postweek = ctq$week - 5

ctq$totfager.cent = ctq$totfager - mean(ctq$totfager)
ctq$age.cent = ctq$age - mean(ctq$age)

# fit multilevel model
M1 = glmer( Y ~  Z  + totfager.cent + postweek + (1 | id ) , family=binomial, data=ctq)
display(M1)

# plot coefficients and associated probabilities
coefficients = coef(M1)$id
coefficients

alpha.hat    = coefficients[,1]
hist(alpha.hat, nclass=12)

p.hat        = exp(alpha.hat) / (1 + exp(alpha.hat))
hist(p.hat, nclass=12)



# Fit morginal models with GEE using different corr structures

G1.indep = geeglm(Y ~ Z + postweek + totfager.cent, family=binomial("logit"), data=ctq, 
                  id=id, corstr="independence", waves=week)
summary(G1.indep)



G1.exch = geeglm(Y ~ Z + postweek + totfager.cent, family=binomial("logit"), data=ctq, 
                  id=id, corstr="exchangeable", waves=week)
summary(G1.exch)


G1.unst = geeglm(Y ~ Z + postweek + totfager.cent, family=binomial("logit"), data=ctq, 
                 id=id, corstr="unstructured", waves=week)
summary(G1.unst)


```

# Question 1

## 1a.) Using the output from Model M1, construct a table with estimates and standard errors for each of the parameters in the model. For the τ parameter, you do not have to write the standard error.

```{r, include = FALSE}

dM1 = display(M1)
sM1 = summary(M1)

fix.1 = data.frame(Estimate = dM1$coef, SE = dM1$se)
ran.1 = data.frame(Tau = sqrt(sM1$varcor$id), SE = NA)
colnames(ran.1) = c('Estimate', 'SE')
table1 = rbind(fix.1, ran.1)
table1 = round(table1, 3)
row.names(table1)[5] = 'Tau'

```

```{r, echo = FALSE}

kable(table1, 'latex', caption = "M1 Sumary", booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  group_rows("Fixed Effects", 1, 4) %>%
  group_rows("Random Effect", 5, 5)

```

## 1b.) Using the model output, provide an estimate of smoking cessation at week 5 for an individual having αi = 0, F = 0, and Z = 0. Do the same for an individual having αi = 0, F = 0, and Z = 1.

```{r, echo = FALSE}

### Prediction for control 
person1 = data.frame(postweek = 0,
                     totfager.cent = 0,
                     Z = 0,
                     id = 'control')

person1$Probability = invlogit(predict(M1, newdata = person1, allow.new.levels = TRUE))


### Prediction for Exercise group
person2 = data.frame(postweek = 0,
                     totfager.cent = 0,
                     Z = 1,
                     id = 'exercise')

person2$Probability = invlogit(predict(M1, newdata = person2, allow.new.levels = TRUE))

one.b = rbind(person1, person2)

kable(one.b, 'latex', caption = "Probability of Cessation for Exercise vs. Control") %>%
  kable_styling(latex_options = c("hold_position"))

```

## 1c.) The R program contains code to construct two histograms, one of the alphai hat and another of a variable pi hat. What quantities are being depicted in each of these histograms? Please be specific with your answer, referring to relevant populations defined by the covariates as appropriate.

$\widehat{\alpha_i}$ are the predicted log odds for each individual if they were to have an average Fagerstrom score and be in the control group at week 5. It can be interpretted as an individual's baseline relative propensity to quit.

$\widehat{p_i}$ are the predicted probabilities of quitting smoking for each individual if they were to have an average Fagerstrom score and be in the control group at week 5. It can be interpretted as an individual's underlying baseline probability of quitting.

## 1d.) Histograms like the ones in the previous question can sometimes be helpful in assessing whether the normality assumption makes sense for the random intercept. Based on these histograms, give your assessment of the validity of the normality assumption, and provide a brief justification.

The histogram for $\widehat{\alpha_i}$ does not exhibit the normal distribution bell shape. This is evidence that our normality assumption is violated. We further see from the predicted probability of smoking cessation histogram that many of the predictions are near 0. The probabilities are heavily right skewed, which is further evidence that our normaility assumption does not hold. 

## 1e.) Provide an interpretation of beta3, the coefficient of Z from this model.

$\beta_3 = 1.883$ is the average increase in log odds of quitting for people in the exercise group compared to the control group, all else equal. $exp(1.883) = 6.57$. On average, the odds of quitting smoking for an individual in the exercise group are 6.57 times higher than in the control group, all else equal. 

## 1f.) Provide an interpretation of the coefficient beta0. 

$\beta_0 = -4.628$ is the average log odds of quitting for people in the control group at week 5 with an average Fagerstrom score.  $logit^{-1}(-4.628) = .01$. The average probability of quitting for an individual in the control group at week 5 with an average Fagerstrom score is .01.

## 1g.) Using the approximation discussed in class, convert the coefficients of Z, F, and t to their "population-averaged" counterparts.

I calculated the population averaged effect, $\beta^*$, using the following formula:

$$  \beta^* = \frac{\beta}{(1 + .346(5.944^2))^{1/2}}$$

```{r, echo = FALSE}

divisor = (1+.346*5.944^2)^.5

one.g = data.frame(Sub_Spec = dM1$coef)
one.g$Pop_Ave = one.g$Sub_Spec/divisor
one.g = one.g[-1,]

kable(one.g, 'latex', 
      row.names = TRUE, booktabs = TRUE, 
      caption = "Subject Specific vs. Population Average Effects") %>%
  kable_styling(latex_options = c("hold_position"))
  

```

# Question 2

## 2a.) Which of the GLM correlation structures in the same as for the multilevel model?

The exchangeable correlation structure corresponds to the random intercept multilevel model.

## 2b.) Construct a table where the row entries are the regression parameters and the scale parameter phi and the columns correspond to each model. Fill the table with the estimate and standard error of each parameter.

```{r, echo = FALSE}

sum.ind = summary(G1.indep)
ind = sum.ind$coefficients[,1:2]
ind2 = sum.ind$dispersion
ind3 = rbind(ind, ind2)

sum.exch = summary(G1.exch)
exch = sum.exch$coefficients[,1:2]
exch2 = sum.exch$dispersion
exch3 = rbind(exch, exch2)


sum.unst = summary(G1.unst)
unst = sum.unst$coefficients[,1:2]
unst2 = sum.unst$dispersion
unst3 = rbind(unst, unst2)

two.b = cbind(ind3, exch3, unst3)
rownames(two.b)[5] = 'phi'
two.b = round(two.b, 3)

kable(two.b, 'latex', row.names = TRUE, caption = "Comparing Correlation Structures", booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position")) %>%
  add_header_above(c(" ", "Independant" = 2, "Exchangeable" = 2, "Unstructured" = 2)) %>%
  group_rows("Coefficients", 1, 4) %>%
  group_rows("Scale Parameter", 5, 5)

```

## 2c.) What is the interpretation of gamma3 in model G1.unst?

$\gamma_3 = .361$ is the average increase in log odds of quitting for the exercise group compared to the control group, all else equal. $exp(.361) = 1.43$. On average, being in the exercise group is associated with 1.43 times the odds of quitting smoking compared to being in the control group, all else equal. 

## 2d.)  Which of these models gives an estimate of gamma3 that is closest to the *marginalized* estimate of beta3 from the multilevel model?

The independent correlation structure is closest to our marginalized estimate from the multilevel model. 

## 2e.) Construct another table where the row entries correspond to estimates of P(Yij = 1) when (a) tj = 0, F = 0 and Z = 0; and (b) tj = 0, F = 0 and Z = 1. The column entries correspond to each model.

```{r, echo = FALSE}

person1$Independent = invlogit(predict(G1.indep, newdata = person1))
person1$Exchangeable = invlogit(predict(G1.exch, newdata = person1))
person1$Unstructured = invlogit(predict(G1.unst, newdata = person1))

person2$Independent = invlogit(predict(G1.indep, newdata = person2))
person2$Exchangeable = invlogit(predict(G1.exch, newdata = person2))
person2$Unstructured = invlogit(predict(G1.unst, newdata = person2))

two.e = rbind(person1, person2)
two.e = two.e %>%
  select(-Probability)

kable(two.e, 'latex', caption = "Predicted Probability of Quitting for Each Cor Structure", booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position")) 

```

## 2f.) Provide an explanation as to why these estimates are different from each other.

## 2g.) If you were reporting the results of this study for a journal, which model(s) would you report and why, and what would be your conclusion about treatment effect?






